Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-03-30 article/02_Materials_and_Methods.Rmd
embed new tables in text

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the time frame of interest was divided into sections of 200 years. For each of the $1400/200 = 7$ section and each of the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section were tested for correlation with the Mantel test, which uses permutations of potentially correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test (Table \ref{tab:distanceandcorrelationtable*} C).

Defining a measure of spatial distance for the above established main units of observation, the macro-regions, is only possible in a highly simplified way. A continuous distance metric such as least-cost path distance cannot be employed on this scale, but it is possible to distinguish ordinal distance classes. For the eight regions, four different categories of spatial distance were defined (Figure \ref{map_regions} B). The test for correlation between spatial and cultural distance works similar to the one between the two cultural distances above, except that the Mantel test has to be applied with Spearman's rank correlation coefficient [@spearmanProofMeasurementAssociation1904] here, because of the different data levels of the measurements (Table \ref{tab:distanceandcorrelationtable*} AB).

```{r distanceandcorrelationtable, echo=FALSE, tidy=FALSE}
load("../analysis/data/output_data/distance_and_correlation_table.RData")
distance_and_correlation_table %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("Mean|Correlation")),
    list(~kableExtra::cell_spec(., "latex", color = ifelse(. < 0, "red", "black"))) 
  ) %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("p-Value")),
    list(~kableExtra::cell_spec(., "latex", bold = . < 0.1)) 
  ) %>%
knitr::kable(
  "latex", 
  caption = "Mean distance, correlation and mantel test results for cultural and spatial distance. Listed are the overall mean squared Euclidian distances for each 200 years time slot of all region-region relations and the result statistics of the Mantel test for the variables \\textit{burial type (bt)} (\\textbf{A}) and \\textit{burial construction (bc)} (\\textbf{B}) with the spatial (sp) distance matrix (Figure \\ref{map_regions} B). \\textbf{C} shows the Mantel test results for the two variables if compared directly. $p$-values below $0.1$ are printed bold.", 
  booktabs = T,
  linesep = "\\addlinespace \\hline \\addlinespace",
  escape = F
) %>%
  kableExtra::add_header_above(
    c(" ", "burial type & spatial distance" = 3, "burial construction & spatial distance" = 3, "variable correlation" = 2)
  ) %>%
  kableExtra::add_header_above(
    c(" ", "A" = 3, "B" = 3, "C" = 2),
    bold = T, line = F
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position"),
    font_size = 9,
    full_width = T
  ) %>%
  kableExtra::column_spec(c(2,4,5,7,9), width = "1.2cm") %>%
  kableExtra::column_spec(c(3,6,8), width = "1.8cm") %>%
  gsub("table", "table*", .)
```

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../analysis/figures/correlation_time_series.jpeg}
\caption{\label{simulation}Development of correlation between cultural and spatial distance. The results for the burial customs data (Table \ref{tab:distanceandcorrelationtable*} AB) are displayed with a solid (\textit{burial type}) and dashed (\textit{burial construction}) line. The horizontal line marks the correlation coefficient of zero, where neither positive nor negative correlation can be assumed.}
\end{figure}

>>>>

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the time frame of interest was divided into sections of 200 years. For each of the $1400/200 = 7$ section and each of the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section were tested for correlation with the Mantel test, which uses permutations of potentially correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test (Table \ref{tab:dacta} C).

Defining a measure of spatial distance for the above established main units of observation, the macro-regions, is only possible in a highly simplified way. A continuous distance metric such as least-cost path distance cannot be employed on this scale, but it is possible to distinguish ordinal distance classes. For the eight regions, four different categories of spatial distance were defined (Figure \ref{map_regions} B). The test for correlation between spatial and cultural distance works similar to the one between the two cultural distances above, except that the Mantel test has to be applied with Spearman's rank correlation coefficient [@spearmanProofMeasurementAssociation1904] here, because of the different data levels of the measurements (Table \ref{tab:dacta} AB).

```{r dacta, echo=FALSE, tidy=FALSE}
load("../analysis/data/output_data/distance_and_correlation_table_A.RData")
distance_and_correlation_table_A %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("Mean|Correlation")),
    list(~kableExtra::cell_spec(., "latex", color = ifelse(. < 0, "red", "black"))) 
  ) %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("p-Value")),
    list(~kableExtra::cell_spec(., "latex", bold = . < 0.1)) 
  ) %>%
knitr::kable(
  "latex", 
  caption = "Mean distance, correlation and mantel test results for cultural and spatial distance. Listed are the overall mean squared Euclidian distances for each 200 years time slot of all region-region relations and the result statistics of the Mantel test for the variables \\textit{burial type (bt)} (\\textbf{A}) and \\textit{burial construction (bc)} (\\textbf{B}) with the spatial (sp) distance matrix (Figure \\ref{map_regions} B). $p$-values below $0.1$ are printed bold.", 
  booktabs = T,
  linesep = "\\addlinespace \\hline \\addlinespace",
  escape = F
) %>%
  kableExtra::add_header_above(
    c(" ", "burial type & spatial distance" = 3, "burial construction & spatial distance" = 3)
  ) %>%
  kableExtra::add_header_above(
    c(" ", "A" = 3, "B" = 3),
    bold = T, line = F
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position"),
    font_size = 9,
    full_width = T
  ) %>%
  kableExtra::column_spec(c(1), width = "4cm") %>%
  gsub("table", "table*", .)
```

```{r dactb, echo=FALSE, tidy=FALSE}
load("../analysis/data/output_data/distance_and_correlation_table_B.RData")
distance_and_correlation_table_B %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("Correlation")),
    list(~kableExtra::cell_spec(., "latex", color = ifelse(. < 0, "red", "black")))
  ) %>%
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("p-Value")),
    list(~kableExtra::cell_spec(., "latex", bold = . < 0.1))
  ) %>%
knitr::kable(
  "latex",
  caption = "Like Table \\ref{tab:dacta}. \\textbf{C} shows the Mantel test results for the two variables if compared directly, \\textbf{D} the results of the Partial Mantel test considering spatial distance as a second independent variable.",
  booktabs = T,
  linesep = "\\addlinespace \\hline \\addlinespace",
  escape = F
) %>%
  kableExtra::add_header_above(
    c(" ", "variable correlation" = 2, "variable correlation with spatial correction" = 2)
  ) %>%
  kableExtra::add_header_above(
    c(" ", "C" = 2, "D" = 2),
    bold = T, line = F
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position"),
    font_size = 9,
    full_width = T
  ) %>%
  kableExtra::column_spec(c(1), width = "4cm") %>%
  gsub("table", "table*", .)
```

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../analysis/figures/correlation_time_series.jpeg}
\caption{\label{simulation}Development of correlation between cultural and spatial distance. The results for the burial customs data (Table \ref{tab:dacta} AB) are displayed with a solid (\textit{burial type}) and dashed (\textit{burial construction}) line. The horizontal line marks the correlation coefficient of zero, where neither positive nor negative correlation can be assumed.}
\end{figure}

### 2019-03-30 article/03_Results.Rmd
further work on the text in reaction to the review: small sample size issue
embed new tables in text

Figure \ref{development_burial_type} and \ref{development_burial_construction} show the region-wise time series of trait occurrence. If we consider these results in isolation, we can construct the following narrative from them: \newline

@@

Table \ref{tab:distanceandcorrelationtable*} C shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity of developments in these regions concerning the dominance of flat or mound graves. Only for the time slot at the transition to the Middle Bronze Age from 1800 to 1600 calBC the 5% significance level of the Mantle test is almost undercut, thus suggesting a statistically significant relationship.

For the relationship between the SED networks and the spatial distance network several observations result from Table \ref{tab:distanceandcorrelationtable*} AB: Cultural distance generally does not seem to correlate with spatial distance. For the variable *burial type* an exception with significant correlation between the distance measures only reveals itself for the Early Bronze Age time slot 2000 to 1800 calBC. In the Early Bronze Age, long before the universal switch to the Urnfield burial tradition, large regions directly adjacent to each other tend to see the same developments. In the Middle Bronze Age the SED developments of the *burial type* variable undergo the peaks already observed. The correlation between cultural and spatial distance during these peaks is low and even tilts between 1400 and 1200 calBC, at the transition to the Late Bronze Age, into the negative -- although not significantly. The cultural distance between several geographically distant regions is then smaller than between several directly adjacent regions. Looking at the SED network of *burial construction*, similar conclusions can be drawn: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated. As well as in *burial type* the Early Bronze Age is by far the phase with the most substantial spatial correlation, though it is only significant for the time slot between 1800 and 1400 calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.

>>>>

Figure \ref{development_burial_type} and \ref{development_burial_construction} show the region-wise time series of trait occurrence. If we consider these results in isolation and also ignore issues of small sample size, we can construct the following narrative from them: \newline

@@

Table \ref{tab:dacta} C shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity of developments in these regions concerning the dominance of flat or mound graves. Only for the time slot at the transition to the Middle Bronze Age from 1800 to 1600 calBC the 5% significance level of the Mantle test is almost undercut, thus suggesting a statistically significant relationship.

For the relationship between the SED networks and the spatial distance network several observations result from Table \ref{tab:dacta} AB: Cultural distance generally does not seem to correlate with spatial distance. For the variable *burial type* an exception with significant correlation between the distance measures only reveals itself for the Early Bronze Age time slot 2000 to 1800 calBC. In the Early Bronze Age, long before the universal switch to the Urnfield burial tradition, large regions directly adjacent to each other tend to see the same developments. In the Middle Bronze Age the SED developments of the *burial type* variable undergo the peaks already observed. The correlation between cultural and spatial distance during these peaks is low and even tilts between 1400 and 1200 calBC, at the transition to the Late Bronze Age, into the negative -- although not significantly. The cultural distance between several geographically distant regions is then smaller than between several directly adjacent regions. Looking at the SED network of *burial construction*, similar conclusions can be drawn: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated. As well as in *burial type* the Early Bronze Age is by far the phase with the most substantial spatial correlation, though it is only significant for the time slot between 1800 and 1400 calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.

### 2019-03-30 article/04_Discussion.Rmd
embed new tables in text

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:distanceandcorrelationtable*}). Only for small periods of time this correlation is visible in the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:

>>>>

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:dacta}). Only for small periods of time this correlation is visible in the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:
